Investigation of the activities performed by experimental researchers in a software engineering lab using an ethnographic methodology

### 2024-06-22 tex/conclusion.tex
Todos los cambios litos, menos la tabla final

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group where we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the lab and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests superficially, e.g., when researchers from different experimental families discuss their mental models.

>>>>

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group where we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the lab and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k,Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental,loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests superficially, e.g., when researchers from different experimental families discuss their mental models.

### 2024-06-22 tex/discussion.tex
Todos los cambios litos, menos la tabla final

Although all empirical researchers know these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature. To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}. Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.

@@

\item The replication problem is related to knowledge transfer \cite{Shull-2004-Knowledge-sharing-issues-SE}. Nevertheless, we wonder what exactly does it mean to transfer knowledge? To what extent or how should it be done? For example, Ferreira \textit{et al.} \cite{Ferreira-2017-planning-experiments} emphasize methodological consistency over discussing methods and techniques.

@@

The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement. Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE}, \cite{Sjoberg-2005-survey-experiments-SE}, \cite{Shull-2004-Knowledge-sharing-issues-SE}. We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character. Instead, we aimed to observe how experimenters work and found out that they perform tasks other than those portrayed in textbooks. \textbf{This is our essential contribution}.

>>>>

Although all empirical researchers know these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature. To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE,Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}. Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.

@@

\item The replication problem is related to knowledge transfer \cite{Shull-2004-Knowledge-sharing-issues-SE,collins2022collective}. Nevertheless, we wonder what exactly does it mean to transfer knowledge? To what extent or how should it be done? For example, Ferreira \textit{et al.} \cite{Ferreira-2017-planning-experiments} emphasize methodological consistency over discussing methods and techniques.

@@

The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement. Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE,Sjoberg-2005-survey-experiments-SE,Shull-2004-Knowledge-sharing-issues-SE}. We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character. Instead, we aimed to observe how experimenters work and found out that they perform tasks other than those portrayed in textbooks. \textbf{This is our essential contribution}.

### 2024-06-22 tex/introduction.tex
Todos los cambios litos, menos la tabla final

\section{Introduction}\label{sec-introduction}
Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives} \cite{Baker-2016-lid-reproducibility}. In Empirical Software Engineering (ESE)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.

Several reasons prevent replication. They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE} \added[id=v3]{\cite{cockburn2020threats}}. In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.

Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments. Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE}, \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.

@@

We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.

>>>>

\section{Introduction}\label{sec-introduction}
Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives,Baker-2016-lid-reproducibility}. In Empirical Software Engineering (ESE)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.

Several reasons prevent replication. They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments,Demagalhaes-2015-replications-SE,cockburn2020threats,mahmood2018reproducibility,dos2022investigating}. In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.

Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments. Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE,Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.

@@

We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE,zhang2019ethnographic} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.

### 2024-06-22 tex/research-execution.tex
Todos los cambios litos, menos la tabla final

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers conducted interviews to clarify the RGUS knowledge and convert it directly into models.

@@

\subsubsection{Replication planning}
Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE}, \cite{Vegas-2006-communication-researchers}, \cite{Miller-2005-replicating-SE-experiments}, \cite{Gomez-2014-understanding-replication}, \cite{Demagalhaes-2015-replications-SE}, \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities. For example, how to approach the experimental design, which experimental objects are necessary, and in which order. We requested the cooperation of the experimenters who ran the baseline experiment to schedule the experimental activities and adapt the design to the new context where the replication would be run.

@@

However, we were not completely satisfied with the ethnographic study. We replicated \textbf{one} experiment in \textbf{one} concrete SE area. Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}. The presence of tacit knowledge represented a severe threat to validity.

@@

We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences. We gave up note-taking and captured the interviews with video, including the experimenters' body language. We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative. We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols}. We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.

@@

The focus groups prompted a substantial growth in the models' details after heated discussions that reconfirmed the existence of terminological diversity \textit{even among experimenters who carried out similar activities within the same family of experiments}. The most likely reason seems to be the training that each experimenter received. Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions. The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches}, \cite{montgomery-2019-Design-Analysis-Experiments}, did not drive self-education. Few of these materials existed when the experimenters started their careers. The experimenters chose the training materials according to their preferences and the experimental families in which they participated. Terminological diversity was the logical consequence.

>>>>

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support,Strandberg-2019-Ethical-Interviews-SE,Yang-2021-interview-study-developers,modi2023using,runeson2020challenges}. Accordingly, the researchers conducted interviews to clarify the RGUS knowledge and convert it directly into models.

@@

\subsubsection{Replication planning}
Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE,Vegas-2006-communication-researchers,Miller-2005-replicating-SE-experiments,Gomez-2014-understanding-replication,Demagalhaes-2015-replications-SE,Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities. For example, how to approach the experimental design, which experimental objects are necessary, and in which order. We requested the cooperation of the experimenters who ran the baseline experiment to schedule the experimental activities and adapt the design to the new context where the replication would be run.

@@

However, we were not completely satisfied with the ethnographic study. We replicated \textbf{one} experiment in \textbf{one} concrete SE area. Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE,Gomez-2014-understanding-replication}. The presence of tacit knowledge represented a severe threat to validity.

@@

We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences. We gave up note-taking and captured the interviews with video, including the experimenters' body language. We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative. We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols,sowden2020verbal}. We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.

@@

The focus groups prompted a substantial growth in the models' details after heated discussions that reconfirmed the existence of terminological diversity \textit{even among experimenters who carried out similar activities within the same family of experiments}. The most likely reason seems to be the training that each experimenter received. Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions. The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches,montgomery-2019-Design-Analysis-Experiments}, did not drive self-education. Few of these materials existed when the experimenters started their careers. The experimenters chose the training materials according to their preferences and the experimental families in which they participated. Terminological diversity was the logical consequence.

### 2024-06-22 tex/research-method.tex
Todos los cambios litos, menos la tabla final

Ethnography is a \textit{participant observation} method \added[id=v2]{used with some frequency in SE, e.g., \cite{traini2022exploring, azarova2022just}}. It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}. Ethnographic researchers adopt "the ESE researchers' point of view" to better understand the reason for their actions \cite[pp. 55-56]{Outhwaite-2007-sage}. Thus, the biases and misperceptions that ESE researchers might have about themselves can be detected. Ethnography also captures the rich context that surrounds ESE researchers.

>>>>

Ethnography is a \textit{participant observation} method \added[id=v2]{used with some frequency in SE, e.g., \cite{traini2022exploring,azarova2022just,de2022practices,wang2024ethnographic}}. It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}. Ethnographic researchers adopt "the ESE researchers' point of view" to better understand the reason for their actions \cite[pp. 55-56]{Outhwaite-2007-sage}. Thus, the biases and misperceptions that ESE researchers might have about themselves can be detected. Ethnography also captures the rich context that surrounds ESE researchers.
